

```{r}
library(tidyverse)
```

```{r}
# Naming & reading all the datasets 
bor_nestlings <- read_csv(here::here("data/73_species.csv")) 
# we use the here::here thing when quarto-ing --- because quorto is in reports and data is somewhere else...

bor_traits <- read_csv(here::here("data/Traits_73_species.csv"))  

avonet <- readxl::read_excel(here::here("data/AVONET Supplementary dataset 1.xlsx"), sheet = "AVONET1_BirdLife")
# has a BIG warning message...? Max says this is ok. 

```



```{r}

# Are nestlings showing up earlier in the year over time?


# nestling data grouped by year, with the mean day of the each year being calculated
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

# Plotting the above mean day for each year with a linear model 
ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")

```


Above ⬆️ combines data for all species. Below ⬇️ breaks down by species

```{r}
# Grouping by both year AND species
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
#fark that is messy...? 
```

Because ⬆️ is chaos, in the next bit ⬇️ we pull out the 5 species with the most data

```{r}
# Figures out which species has the most data
data_richness <- bor_nestlings %>% 
  count(Species)

# From ChatGPT: This line uses the arrange function to reorder the rows of the data frame based on a specific      column. In this case, it is ordering the data frame by the column n in descending order. 
# desc is short for descending, which means the data will be sorted from highest to lowest.
most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% #filters by species then by the species with the most data
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The above model has a general pattern, with one conflicting example. 
What species have the strongest trends?

```{r}
# I'm giving you a *function* to help here. You'll learn more about those next week.

# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))

```

----------------- ⬆️ YOU LEFT OFF THERE ⬆️--------------------------------------

Spot check two species:
```{r}
soi <- c("ARDCIN", "LARMIN") #species of interest

species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```


# ------------------------------------------------------------------------------
                                  YOUR TURN
# ------------------------------------------------------------------------------

Find the species with the most positive and most negative trends, then plot just those.

```{r}
#| eval: false
nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  # Sort by the day of year trend
  ___(___) %>% 
  # Keep just the first (most negative trend) and last (most positive trend) rows
  slice(c(___, ___))

# Now plot them
bor_by_year %>% 
  filter(Species %in% ___) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

 ------------------- Bring in more data --------------------
 
Next we’re going to test the hypothesis that smaller birds have more flexible phenology
(i.e. the absolute value of the trend is correlated with smaller body sizes.)

To test our hypothesis, we need to add AVONET data to our phenology data by joining. 
The keys are a bit tricky here!

   -- bor_nestlings has a column called Species with a 6-letter code.
 
   -- bor_traits has a column called Abbreviation with the 6-letter code 
    and a column called `Scientific name` with the binomial name.

   -- avonet has column called Species1 with the binomial name.

Mission: We need to JOIN bor_nestlings to bor_traits, then JOIN with avonet.

```{r}
#| eval: false
# First = select and rename important columns: 
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)

bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = `Scientific name`)
```

Now Join:: 
```{r}
#| eval: false
bor_trends_size <- bor_trends %>% 
  left_join(___, by = ___) %>% 
  left_join(___, by = ___) %>% 
  mutate(abs_trend = abs(doy_trend))

# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")

```

That’s  ⬆️ kind of hard to interpret.
Make a new variable: bor_trends_size2
  this removes the missing values in Mass_g and keeps only birds smaller than 2000g 
  with an absolute trend less than 1.
  
  
```{r}
#| eval: false
bor_trends_size2 <- ???
  
ggplot(bor_trends_size2, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Is our hypothesis supported?

------------
Submission
Render index.qmd, which produces reports/index.html. Move your HTML file to docs/.
Enable GitHub pages for your repo.
Commit your changes and push.
Open an issue and tag me in it.
